<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{sotkanet R vignette}
-->

Sotkanet API R tools
===========

This is the [sotkanet](https://github.com/rOpenGov/sotkanet) R package to access data from the [Sotkanet portal](http://uusi.sotkanet.fi/portal/page/portal/etusivu/hakusivu) that provides over 2000 demographic indicators across Finland and Europe, maintained by the National Institute for Health and Welfare (THL). For more information, see [Sotkanet indicator database](http://uusi.sotkanet.fi/portal/page/portal/etusivu/tietoa_palvelusta) and [API description](http://uusi.sotkanet.fi/portal/pls/portal/!PORTAL.wwpob_page.show?_docname=22001.PDF). This package is part of [rOpenGov](http://ropengov.github.com/).


```{r, echo = FALSE}
# HW preferred defaults for displaying code
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

### Installation

Release version

```{r install, eval=FALSE}
install.packages("sotkanet")
```


Development version (for the adventurous ones)

```{r install2, eval=FALSE}
library(devtools)
install_github("ropengov/sotkanet")
```



### Listing available indicators

Load sotkanet and other libraries used in this vignette

```{r loadlib, warning=FALSE, message=FALSE, echo=FALSE, eval=TRUE}
library(sotkanet)
library(knitr) 
```

List available Sotkanet indicators:

```{r sotkanetIndicators, warning=FALSE, message=FALSE, eval=FALSE}
sotkanet.indicators <- SotkanetIndicators(type = "table")
# kable(head(sotkanet.indicators))
```

List geographical regions with available indicators:

```{r sotkanetRegions, warning=FALSE, message=FALSE, eval=TRUE}
sotkanet.regions <- SotkanetRegions(type = "table")
kable(head(sotkanet.regions))
```


### Querying SOTKAnet indicators

Get the [indicator 10013](http://uusi.sotkanet.fi/portal/page/portal/etusivu/hakusivu/metadata?type=I&indicator=10013) from Finland (Suomi) for 1990-2012 (Eurostat employment statistics youth unemployment), and plot a graph:

```{r sotkanetData, warning=FALSE, message=FALSE, eval=TRUE}
# Get indicator data
dat <- GetDataSotkanet(indicators = 10013, years = 1990:2012, 
       		       genders = c('female', 'male', 'total'), 
		       region.category = "EUROOPPA", regions = "Suomi")

# Investigate the first lines in the data
kable(head(dat))
```

### Fetch all SOTKAnet indicators

This takes for a long time and is not recommended for regular
use. Save the data on your local disk for further work.

```{r sotkanetDataAll, warning=FALSE, message=FALSE, eval=FALSE}
# These indicators have problems with R routines:
probematic.indicators <- c(1575, 1743, 1826, 1861, 1882, 1924, 1952, 2000, 2001, 2033, 2050, 3386, 3443)

# Get data for all indicators
datlist <- list()
for (ind in setdiff(sotkanet.indicators$indicator, probematic.indicators)) {
  datlist[[as.character(ind)]] <- GetDataSotkanet(indicators = ind, 
  		years = 1990:2013, genders = c('female', 'male', 'total'))
}

# Combine tables (this may require considerable time and memory 
# for the full data set)
dat <- do.call("rbind", datlist)
```

## Visualization

Download and visualize time series:

```{r sotkanetDataVisu, warning=FALSE, message=FALSE, fig.width=10, figh.height=5, eval=TRUE}
# Pick indicator name
indicator.name <- as.character(unique(dat$indicator.title.fi))
indicator.source <- as.character(unique(dat$indicator.organization.title.fi))

# Visualize
library(ggplot2)
theme_set(theme_bw(20)); 
p <- ggplot(dat, aes(x = year, y = primary.value, group = gender, color = gender)) 
p <- p + geom_line() + ggtitle(paste(indicator.name, indicator.source, sep = " / ")) 
p <- p + xlab("Year") + ylab("Value") 
p <- p + theme(title = element_text(size = 10))
p <- p + theme(axis.title.x = element_text(size = 20))
p <- p + theme(axis.title.y = element_text(size = 20))
p <- p + theme(legend.title = element_text(size = 15))
print(p)
```


Investigate the effect of municipality size on demographic
variation. Smaller municipalities show more random variation as
expected by statistical arguments:

```{r sotkanetVisu3, warning=FALSE, message=FALSE, eval=TRUE, fig.width=10, fig.height=5}
selected.inds <- c(127, 178)
dat <- GetDataSotkanet(indicators = selected.inds, 
       			years = 2011, genders = c('total'))
# Pick necessary fields and remove duplicates
datf <- dat[, c("region.title.fi", "indicator.title.fi", "primary.value")]
datf <- datf[!duplicated(datf),]
dw <- reshape(datf, idvar = "region.title.fi", 
      		    timevar = "indicator.title.fi", direction = "wide")
names(dw) <- c("Municipality", "Population", "Migration")
p <- ggplot(dw, aes(x = log10(Population), y = Migration)) +
       geom_point(size = 3)
       ggtitle("Migration vs. population size") 
       theme(title = element_text(size = 15))
       theme(axis.title.x = element_text(size = 20))
       theme(axis.title.y = element_text(size = 20))
       theme(legend.title = element_text(size = 15))
print(p)
```

## Further examples

For further usage examples, see
[Louhos-blog](http://louhos.wordpress.com), and
[takomo](https://github.com/louhos/takomo/tree/master/Sotkanet), and 
[Helsinki Region Infoshare-blog](http://www.hri.fi/fi/ajankohtaista/sotkanet-indikaattoripankki-mukaan-sorvi-tyokalupakkiin/)


# Licensing and Citations

### SOTKAnet data

Cite SOTKAnet and link to [http://www.sotkanet.fi](http://www.sotkanet.fi/). Also mention indicator provider. 

 * [Full license and terms of use](http://uusi.sotkanet.fi/portal/page/portal/etusivu/tietoa_palvelusta). 

Central points:

 * SOTKAnet REST API is meant for non-regular data queries. Avoid regular and repeated downloads.
 * SOTKAnet API can be used as the basis for other systems
 * Metadata for regions and indicators are under [CC-BY 3.0](http://creativecommons.org/licenses/by/3.0/) 
 * THL indicators are under [CC-BY 3.0](http://creativecommons.org/licenses/by/3.0/) 
 * Indicators provided by third parties can be used only by separate agreement!


### SOTKAnet R package

This work can be freely used, modified and distributed under the 
[Two-clause BSD license](http://en.wikipedia.org/wiki/BSD\_licenses).

```{r citation, message=FALSE, eval=TRUE}
citation("sotkanet")
```

### Session info

This vignette was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessionInfo()
```
