---
title: "Sotkanet API R tools"
author: "Leo Lahti"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sotkanet API R tools}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: inline
---

This is the [sotkanet](https://github.com/rOpenGov/sotkanet) R package to access data from the [Sotkanet portal](https://sotkanet.fi/sotkanet/en/index) that provides over 2000 demographic indicators across Finland and Europe, maintained by the National Institute for Health and Welfare (THL). For more information, see [Information about Sotkanet](https://sotkanet.fi/sotkanet/en/tietoa-palvelusta) and [API description](https://sotkanet.fi/sotkanet/en/ohje/74). This package is part of [rOpenGov](http://ropengov.org).

```{r, echo = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  purl = NOT_CRAN,
  eval = NOT_CRAN
)
```

## Installation

Release version from CRAN:

```{r install, eval=FALSE}
install.packages("sotkanet")
```

Development version from GitHub:

```{r install2, eval=FALSE}
library(devtools)
install_github("ropengov/sotkanet")
```

## Usage

### Listing available indicators

Load sotkanet and other libraries used in this vignette

```{r loadlib, warning=FALSE, message=FALSE, echo=FALSE}
library(sotkanet)
library(knitr) 
library(magrittr)
library(kableExtra)
```

List available Sotkanet indicators:

```{r sotkanetIndicators, warning=FALSE, message=FALSE}
# Using a pre-defined list of indicators for a smaller download
sotkanet.indicators <- SotkanetIndicators(id = c(4,5,6,7,46,74), type = "table")
kable(head(sotkanet.indicators))
```

List geographical regions with available indicators:

```{r sotkanetRegions, warning=FALSE, message=FALSE}
# options(width = 60)
sotkanet.regions <- SotkanetRegions(type = "table")
kable(head(sotkanet.regions))
```

### Querying SOTKAnet indicators

Get the indicator no. 10013 (Unemployment rate of young people aged 16-24 years) from Sotkanet, from Finland (Suomi) for 1990-2012 (Eurostat employment statistics youth unemployment), and plot a graph:

```{r sotkanetData, warning=FALSE, message=FALSE}
# Get indicator data
dat <- GetDataSotkanet(indicators = 10013, years = 1990:2012, 
       		       genders = c('female', 'male', 'total'), 
		       region.category = "EUROOPPA", regions = "Suomi")

# Investigate the first lines in the data
kable(head(dat)) %>% 
  kable_styling() %>%
  kableExtra::scroll_box(width = "100%")
```

### Visualization

Download and visualize time series:

```{r sotkanetDataVisu, warning=FALSE, message=FALSE, fig.width=10, figh.height=5, out.width = "100%"}
# Pick indicator name
indicator.name <- as.character(unique(dat$indicator.title.fi))
indicator.source <- as.character(unique(dat$indicator.organization.title.fi))

# Visualize
library(ggplot2)
theme_set(theme_bw(20)); 
p <- ggplot(dat, aes(x = year, y = primary.value, group = gender, color = gender)) 
p <- p + geom_line() + ggtitle(paste(indicator.name, indicator.source, sep = " / ")) 
p <- p + labs(x = "Year", y = "Value", caption = "Data source: https://sotkanet.fi")
p <- p + theme(title = element_text(size = 10))
p <- p + theme(axis.title.x = element_text(size = 20))
p <- p + theme(axis.title.y = element_text(size = 20))
p <- p + theme(legend.title = element_text(size = 15))
print(p)
```

Investigate the effect of municipality size on demographic
variation. Smaller municipalities show more random variation as
expected by statistical arguments:

```{r sotkanetVisu3, warning=FALSE, message=FALSE, fig.width=10, fig.height=5, out.width = "100%"}
selected.inds <- c(127, 178)
dat <- GetDataSotkanet(indicators = selected.inds, 
       			years = 2011, genders = c('total'))
# Pick necessary fields and remove duplicates
datf <- dat[, c("region.title.fi", "indicator.title.fi", "primary.value")]
datf <- datf[!duplicated(datf),]
dw <- reshape(datf, idvar = "region.title.fi", 
      		    timevar = "indicator.title.fi", direction = "wide")
names(dw) <- c("Municipality", "Population", "Migration")
p <- ggplot(dw, aes(x = log10(Population), y = Migration)) +
       geom_point(size = 3)
       ggtitle("Migration vs. population size") 
       theme(title = element_text(size = 15))
       theme(axis.title.x = element_text(size = 20))
       theme(axis.title.y = element_text(size = 20))
       theme(legend.title = element_text(size = 15))
print(p)
```

### Fetch all SOTKAnet indicators

This takes for a long time and is not recommended for regular
use. Save the data on your local disk for further work.

```{r sotkanetDataAll, warning=FALSE, message=FALSE, eval=FALSE}
# These indicators have problems with R routines:
problematic.indicators <- c(1575, 1743, 1826, 1861, 1882, 1924, 1952, 2000, 
                            2001, 2033, 2050, 3386, 3443)

# Get data for all indicators
datlist <- list()
for (ind in setdiff(sotkanet.indicators$indicator, problematic.indicators)) {
  datlist[[as.character(ind)]] <- GetDataSotkanet(indicators = ind, 
  		years = 1990:2013, genders = c('female', 'male', 'total'))
}

# Combine tables (this may require considerable time and memory 
# for the full data set)
dat <- do.call("rbind", datlist)
```

### Further examples

For further usage examples, see
[rOpenGov-blog](http://ropengov.org), and
[takomo](https://github.com/louhos/takomo/tree/master/Sotkanet), and 
[Helsinki Region Infoshare-blog](http://www.hri.fi/fi/ajankohtaista/sotkanet-indikaattoripankki-mukaan-sorvi-tyokalupakkiin/)

## Licensing and Citations

### SOTKAnet data

Cite SOTKAnet and link to [https://sotkanet.fi/sotkanet/fi/index](https://sotkanet.fi/sotkanet/fi/index). Also mention indicator provider. 

 * [Full license and terms of use](https://sotkanet.fi/sotkanet/en/tietoa-palvelusta). 

Central points:

 * SOTKAnet REST API is meant for non-regular data queries. Avoid regular and repeated downloads.
 * SOTKAnet API can be used as the basis for other systems
 * Metadata for regions and indicators are under [CC-BY 4.0](https://creativecommons.org/licenses/by/4.0/) 
 * THL indicators are under [CC-BY 4.0](https://creativecommons.org/licenses/by/4.0/) 
 * Indicators provided by third parties can be used only by separate agreement!


### SOTKAnet R package

This work can be freely used, modified and distributed under the 
[Two-clause BSD license](https://en.wikipedia.org/wiki/BSD\_licenses).

```{r citation, message=FALSE, eval=TRUE}
citation("sotkanet")
```

### Session info

This vignette was created with

```{r sessioninfo, message=FALSE, warning=FALSE}
sessionInfo()
```
